Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-08-08 volunteer-clouds-ieee-ci-si.tex
More about the algorithm
This is a brain dump just to capture the ideas. A heavy edit is
expected.

JavaScript has commonly been used to develop client-side scripts that handle the user interface,
communicate asynchronously with servers and update the content that is displayed \cite{flanagan2006javascript}.
Long-running scripts like those proposed in this work can interfere with the responsiveness of 
the web application, because everything is handled by a single thread. 
The Web Workers specification \cite{hickson2012web} defines an API that allows Web developers 
to execute long-running scripts in parallel with the scripts of the main page. 
Worker scripts are spawned in the background allowing a thread-like operation using 
message-passing as the communication mechanism between the worker's execution environment and the main thread. 
According to the Web Workers W3C Candidate Recomendation \cite{hickson2012web} 
workers are expected to be long-lived, they have a high start-up performance cost, 
and a high per-instance memory cost. The Web Worker API is implemented in current
versions of both desktop and mobile web browsers. For volunteer computing using the Web Worker API   
brings several advantages over a single threaded implementation:
\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script 
keeps running in the background even it the user brings forward another tabbed window.
\item Several instances of the NodEO script can be executed in parallel in a single web page. 
\item Many immplemantations of the Web Worker API can use multiple CPUs for the execution
of worker scripts, in particular the execution in multiple CPUs was tested in desktop verions of
Chrome, FireFox and Safari.

\end{itemize}

\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clientts use WebWorkers to run the evolutionary algorithm,
  with two of them per browser.}  
\label{fig:system:w2}
\end{figure}

>>>>

JavaScript has commonly been used to develop client-side scripts
that handle the user interface, communicate asynchronously with servers and 
update the content that is displayed \cite{flanagan2006javascript}.
Long-running scripts like those proposed in this work can interfere with the
responsiveness of the web application, because everything is handled by a 
single thread. The Web Workers specification \cite{hickson2012web} defines an 
API that allows Web developers to execute long-running scripts in parallel
with the scripts of the main page. Worker scripts are spawned in the 
background allowing a thread-like operation using message-passing as the 
communication mechanism between the worker's execution environment and the 
main thread. According to the Web Workers W3C Candidate Recomendation 
\cite{hickson2012web} workers are expected to be long-lived, they have a high 
start-up performance cost, and a high per-instance memory cost. The Web Worker
API is implemented in current versions of both desktop and mobile web browsers.
For volunteer computing using the Web Worker API brings several advantages 
over a single threaded implementation:

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script 
keeps running in the background even it the user brings forward another tabbed
window.
\item Several instances of NodEO can be executed in parallel in a single web
page. 
\item Many immplemantations of the Web Worker API can use multiple CPUs for
the execution of worker scripts, in particular the execution in multiple CPUs
was tested in desktop verions of Chrome, FireFox and Safari.
\item Each NodEO instance can be restarted independenly.
\end{itemize}

A sequence diagram of the W$^2$ version of {\sf NodIO} is presented 
in Figure~\ref{fig:system:w2}. In this version, clients use the Web Workers
API to run two NodEO instances per browser. The sequence diagram shows the 
interaction between processess in a time sequence. In this system there are
two kinds of asynchronus messages: HTTP Requests shown as bold arrows between 
the client and the server where PUT and GET HTTP Methods are represented by 
red and blue arrows respectively. There are also messages between workers and 
the main thread in locally in the client, this are shown as thin solid arrows.
There are three processes involved:
\begin{itemize}
\item {\em nodejs} This is the server side process, implented in the nodejs 
framework. Web request handlers are implemented using the Express web 
framework. For keeping a log of the execution the Winston library is used, and 
finally the NodeIO framework is responsible for keeping the shared pool.
\item {\em Client main thread}  This process handles the user interaction
using the JQuery framwork. The Chart.js library is used to display a chart of
the best fitness in the population over time. The main script also creates 
worker instances and handles the user interface.
\item {\em Worker global scope} The worker execution space does not have 
access to the Document Object Model (DOM) or JQuery objects in the main thread.
Asynchronus HTTP requests to the server are implemented using the XHR object. 
\end{itemize}

The sequence of events are:
\begin{enumerate}
\item First the NodIO process is started in nodejs. As part of the 
initialization procedure the number of experiment is initialised and a log 
file is created. The shared pool implemented as an array is reset.
\item A volunteer follows the link of the experiment triggering a GET HTTP 
request and the page is loaded.
\item The main script renders the page and creates the worker instance. In 
order to create a dedicated worker instance another GET request is sent to the 
server in order to recieve the worker implementation script. 
Each worker instance executes one or more evolutionary algorithms 
{\em islands}. Each starts with random population and it is assigned a 
universally unique  identifier (UUID). The UUID is included in each of the HTTP
requests sent to the server.  
The initial paramaters for each island
can be distinct. Once the worker is created and the parameters have been set,
a message is sent to the worker to start the algorithm.
\item The worker first creates the population following with the previously 
specified parameters. Once the population has been initialised, the EA 
generation loop is started. After n generations the state of the population 
is evaluated. First the chromosome with the best fitness is sent to the shared
population with a PUT HTTP request. If the best fitness is not a global best,
the EA loop must continue, but first a chromosome is randomly selected from 
the shared pupulation. Also the current generation number and the best 
chromosome is sent to the client main thread in order to update the DOM to 
display the current state of the island (described next). These messages are
all asynchronous so even before a message is recieved by target the EA loop of
the worker could continue. If the best fitness is a global best the best 
chromosme is sent to the server and the loop ends. A message is sent to the 
main thread indicating that the best individual has been found.
\item The main thread recieves messages from workers as events handled by 
a callback function. When an iteration message is recived ,the page is 
updated by displaying a dinamic plot of the geneartion number and best 
fitness, and a representation of the chromosme is also generated.   
\item When a global best is recieved from an island the current experiment 
ends, the experiment number is incremented and the population array is reset.  
\item When the main thread recieves a message the a worker has ended the EA
loop, that worker is reinitialized. The worker process is not ended, to
avoid the cost, only the parameters and population is reset and a new UIID 
assigned.
\end{enumerate}

\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the W$^2$ version of {\sf NodIO}. In this
  version, clients use Web Workers to run the evolutionary algorithm,
  with two of them per browser.}  
\label{fig:system:w2}
\end{figure}
